---
title: "merge_datasets"
author: "Kyle Abraham Campbell"
date: "2023-11-13"
output: html_document
---

```{r setup, include=FALSE}
library(here)
library(minfi)
library(Seurat)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Merge all datasets for deconvolution reference
```{r}
# Common covariates between the datasets
clin.covariates <- c("Sample.ID", "Sex", "CellType", "Case_ID", "Study", "Sentrix_ID", "Sentrix_Position"
)

kc <- readRDS(here("data", "analytic", "campbell", paste0("qc_beta_bmiq_campbell_2023-11-13.rda")))
kc.pd <- readRDS(here("data", "analytic", "campbell", "pd_kc_harmonized_2023-11-13.rda"))
colnames(kc) <- kc.pd$Sample.ID


vy <- readRDS(here("data", "analytic", "yuan", paste0("beta_bmiq_qced_yuan_2023-11-14.rda")))
vy.pd <- readRDS(here("data", "analytic", "yuan", paste0("beta_bmiq_qced_yuan_pd_2023-11-14.rda"))) %>% 
  dplyr::select(clin.covariates)
colnames(vy) <- vy.pd$Sample.ID

blood <- readRDS(here("data", "analytic", "blood", paste0("qc_beta_bmiq_blood_2023-11-16.rda")))
blood.pd <- readRDS(here("data", "analytic", "blood", paste0("qc_pd_blood_2023-11-16.rda")))
colnames(blood) <- blood.pd$Sample.ID

nrbc.bakulski <- readRDS(here("data", "analytic", "nrbc", "beta_noob_bmiq_nrbc_qc_Bakulski_2023-11-15.rda"))
nrbc.bakulski.pd <-readRDS(here("data", "analytic", "nrbc", "bakulski_nrbc_pd_2023-11-16.rda")) %>%
  dplyr::select(clin.covariates)
colnames(nrbc.bakulski) <- nrbc.bakulski.pd$Sample.ID

nrbc.degoede <- readRDS(file = here("data", "analytic", "nrbc", "beta_noob_bmiq_nrbc_qc_deGoede_2023-11-15.rda"))
nrbc.degoede.pd <- readRDS(file = here("data", "analytic", "nrbc", "deGoede_nrbc_pd_2023-11-16.rda")) %>%
  dplyr::select(clin.covariates) %>%
  mutate(Study = "de Goede")
colnames(nrbc.degoede) <- nrbc.degoede.pd$Sample.ID
```

```{r}
pd <- rbind(kc.pd, vy.pd, blood.pd, nrbc.bakulski.pd, nrbc.degoede.pd)
rownames(pd) <- pd$Sample.ID
```


## Find common probes
368435 common probes between all datasets
```{r}
common.probes <- Reduce(intersect, list(rownames(kc), rownames(vy), rownames(blood), rownames(nrbc.bakulski), rownames(nrbc.degoede)))
length(common.probes)
```

## Concatenate
```{r}
beta.list <- 
  list(
    kc, vy, blood, nrbc.bakulski, nrbc.degoede
  )
# Subset each to common probes
beta.list.overlap <- map(beta.list, .f = ~ .x[rownames(.x) %in% common.probes,])

beta.concat <- do.call("cbind", beta.list.overlap)
```

Exploratory CpG plots; an example showing the 4 ST samples we kept appear very specific, in addition to the stromal we left out. cg01802953 was very low in CT/ST and around those tight cluster in nRBC/stromal in the final analytic sample
```{r}
plotCpg(dat = beta.concat, pheno = pd$CellType, type = "categorical", cpg = "cg01802953")
```


## Subset to keep based on placental cell type clustering 
```{r}
keep.ids <- readRDS(here("data", "analytic", "campbell_yuan_concat", "strict_keep_2023-11-15.rda"))

drop.ids <- pd %>%
  filter(Study %in% c("Campbell", "Yuan")) %>%
  filter(!Sample.ID %in% keep.ids) %>%
  pull(Sample.ID)
length(drop.ids)

pd.keep <- pd %>%
  filter(!Sample.ID %in% drop.ids)

pd.drop <- pd %>%
  filter(Sample.ID %in% drop.ids)

beta.concat.keep <- beta.concat[,colnames(beta.concat) %in% pd.keep$Sample.ID]

levels(pd.keep$CellType) <- c("Whole tissue", "B cell", "CD4T", "CD8T", "Cytotrophoblast", "Endothelial", 
"Hofbauer", "Monocyte", "Neutrophil", "Natural Killer", "Nucleated red blood cell", "Stromal", "Syncytiotrophoblast"
)
```

Take a look at excluded samples
```{r}
pd.dropped <- pd %>%
  filter(Study %in% c("Campbell", "Yuan")) %>%
  filter(!Sample.ID %in% keep.ids)
```


## Save
```{r, eval = F}
saveRDS(pd.keep, here("data", "analytic", "concat", "pd_2023-11-30.rda"))
saveRDS(beta.concat.keep, here("data", "analytic", "concat", "beta_2023-11-16.rda"))
```

```{r}
pd <- readRDS(here("data", "analytic", "concat", "pd_2023-11-30.rda"))
beta <- readRDS(here("data", "analytic", "concat", "beta_2023-11-16.rda"))
```

## Visualize

Seurat pipeline
```{r}
pipe.fast.seu <- function(seu) {
  seu <- ScaleData(seu, do.scale = F, do.center = F)
  seu <- RunPCA(seu, features = rownames(seu))
  seu <- FindNeighbors(object = seu)
  seu <- FindClusters(object = seu)
  seu <- RunUMAP(seu, dims = 1:30)
  return(seu)
}
```

```{r}
seu <- CreateSeuratObject(counts = beta, meta.data = pd, data = beta)

# Not sure why the data step is there
#seu <- CreateSeuratObject(counts = beta, data = beta.placenta, meta.data = pd)
```

```{r}
seu <- pipe.fast.seu(seu)
```

Get PCA variance explained
```{r}
pca = seu@reductions$pca
eigValues = (pca@stdev)^2  ## EigenValues
varExplained = eigValues / sum(eigValues)
print(varExplained[1:5])
```

```{r}
levels(seu@meta.data$CellType) <- c("B cell", "CD4T", "CD8T", "Cytotrophoblast", "Endothelial", "Hofbauer", "Monocyte", "Neutrophil", "Natural Killer", "nRBC", "Stromal", "Syncytiotrophoblast", "Whole tissue")
seu.dat <- seu@meta.data

#seu.dat$CellType <- factor(seu.dat$CellType, levels = c("Cytotrophoblast", "Endothelial", "Hofbauer", "Stromal", "Syncytiotrophoblast", "Villi"))
seu.dat$UMAP_1 <- seu@reductions$umap@cell.embeddings %>% as.data.frame() %>% pull(umap_1)
seu.dat$UMAP_2 <- seu@reductions$umap@cell.embeddings %>% as.data.frame() %>% pull(umap_2)
seu.dat$PC_1 <- seu@reductions$pca@cell.embeddings %>% as.data.frame() %>% pull(PC_1)
seu.dat$PC_2 <- seu@reductions$pca@cell.embeddings %>% as.data.frame() %>% pull(PC_2)
```

```{r}
DimPlot(seu, label = T, repel = T, reduction = "pca", group.by = "CellType") + NoLegend()
```

```{r}
levels(seu@meta.data$CellType)
```

```{r}
seu@meta.data$CellType <- factor(seu@meta.data$CellType)
levels(seu@meta.data$CellType)
levels(seu@meta.data$CellType) <- c("B cell", "CD4T", "CD8T", "Cytotrophoblast", "Endothelial", "Hofbauer", "Monocyte", "Neutrophil", "Natural Killer", "Nucleated red blood cell", "Stromal", "Syncytiotrophoblast", "Whole tissue")
```

```{r}
colors.13 <- c(
"#8DD3C7", "#FFFFB3", "#BEBADA",
"#FB8072", "#80B1D3", "#FDB462", 
"#B3DE69", "#FCCDE5", "#FDB462",
"#BC80BD", "#CCEBC5", "#8DD3C7",
"#BC80BD"
)
custom_palette <- colorRampPalette(brewer.pal(12, "Set3"))(13)
custom_palette

palette_13 <- c(brewer.pal(9, "Set1"), brewer.pal(4, "Paired"))
palette_13 %>% dput

custom_palette_13 <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#B2B223", 
"#A65628", "#F781BF", "#999999", "#A6CEE3", "#1F78B4", "#FF7F7F", 
"#33A02C")
```

```{r, eval = F}
DimPlot(seu, label = T, repel = T, reduction = "umap", group.by = "CellType", cols = palette_13, pt.size = 1.5) + ggtitle("") + theme_bw(base_size = 14) + xlab("UMAP 1") + ylab("UMAP 2")  + NoLegend()

seu.dat
```

```{r}
label_data <- seu.dat %>%
  group_by(CellType) %>%
  summarise(
    UMAP_1 = mean(UMAP_1),
    UMAP_2 = mean(UMAP_2)
  )
label_data_custom <- label_data %>%
  mutate(UMAP_1 = ifelse(CellType == "Hofbauer", UMAP_1 - 1, UMAP_1)) %>%
  mutate(UMAP_1 = ifelse(CellType == "Cytotrophoblast", UMAP_1 + 0.5, UMAP_1)) %>%
  mutate(UMAP_2 = ifelse(CellType == "Whole tissue", UMAP_2 + 1, UMAP_2)) %>%
  #mutate(UMAP_2 = ifelse(CellType == "Stromal", UMAP_2 - 0.5, UMAP_2)) %>%
  mutate(UMAP_2 = ifelse(CellType == "Syncytiotrophoblast", UMAP_2 - 1, UMAP_2))


# Basic ggplot with UMAP coordinates and cluster labels
ggplot(seu.dat, aes(x = UMAP_1, y = UMAP_2, color = CellType)) +
  geom_point(size = 1.5, alpha = 0.5) +  # Adjust point size and transparency as needed
  scale_color_manual(values = custom_palette_13) +
  theme_bw() +
  theme(legend.position = "none") +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
   geom_text_repel(
    data = label_data_custom,
    aes(x = UMAP_1, y = UMAP_2, label = CellType),
    size = 5,
    box.padding = 0.5,
    point.padding = 0.5,
    segment.color = "black"
  )

#ggsave(filename = here("results", "analytic", "umap_viz", "umap_all_20241103.tiff"), dpi = 300)
```


```{r, eval =F}
ggsave(filename = here("results", "analytic", "umap_viz", "umap_all_20231130.png"), dpi = 300)
```

```{r}
levels(seu@meta.data$CellType)
```
