---
title: "Salas et al QC"
author: "Kyle Abraham Campbell"
date: '2022-09-07'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# CRAN packages
library(BiocManager)
library(ChAMP)
library(corrplot)
library(data.table)
library(FactoMineR)
library(factoextra)
library(GGally)
library(GEOquery)
library(ggrepel)
library(here)
library(readxl)
library(svd)
library(tidymodels)
library(tidyverse)
library(viridis)

# BioConductor packages
library(ExperimentHub)
library(illuminaio)
library(IlluminaHumanMethylationEPICmanifest)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(minfi)
#library(wateRmelon)

# Github packages
#devtools::install_github("hhhh5/ewastools@master")
library(ewastools)
#devtools::install_github('wvictor14/planet')
library(planet)
#remotes::install_github("immunomethylomics/FlowSorted.Blood.EPIC")
library(FlowSorted.Blood.EPIC)
#remotes::install_github("immunomethylomics/FlowSorted.BloodExtended.EPIC")
library(FlowSorted.BloodExtended.EPIC)
```

# FlowSorted.Blood.EPIC Illumina QC w/ ewastools

```{r, eval = F, include = F}
#blood <- getGEO("GSE110554", GSEMatrix = TRUE, destdir = here("data", "blood_flow_epic"))
#saveRDS(blood, here("data", "blood_flow_epic", "GSE110554_GSE_matrix.rda"))
```

```{r, eval = F}
FlowSorted.Blood.EPIC <-
    libraryDataGet("FlowSorted.Blood.EPIC")

FlowSorted.BloodExtended.EPIC <-
    libraryDataGet("FlowSorted.BloodExtended.EPIC")
```

```{r}
#FlowSorted.Blood.EPIC.pd <- pData(FlowSorted.Blood.EPIC) %>% as.data.frame()
#saveRDS(FlowSorted.Blood.EPIC.pd, here("data", "blood_flow_epic", "pd.rda"))
blood.pd <- readRDS(here("data", "blood_flow_epic", "pd.rda"))
blood.pd$Basename <- paste0(here("data", "blood_flow_epic", "idats"), "/", blood.pd$Array_well)
blood.pd <- blood.pd %>%
  dplyr::rename(
                Sentrix_ID = "Slide",
                Sentrix_Position = "Array")
#saveRDS(blood.pd, here("data", "blood_flow_epic", "pd_formatted.rda"))
table(blood.pd$CellType, blood.pd$Sex)


#Load ewastools manifest
manifest <- data.table::copy(ewastools:::manifest_epic)
```

Not doing extended; likely too many celltypes for this application
```{r, eval = F, include = F}
#FlowSorted.BloodExtended.EPIC.pd <- pData(FlowSorted.BloodExtended.EPIC) %>% as.data.frame()
#saveRDS(FlowSorted.BloodExtended.EPIC.pd, here("data", "bloodextended_flow_epic", "pd.rda"))
blood.ext.pd <- readRDS(here("data", "bloodextended_flow_epic", "pd.rda"))
table(blood.ext.pd$CellType, blood.ext.pd$Sex)
```

Read in blood dataset and add metadata
```{r, eval = F}
blood <- read_idats(paste0(here("data", "blood_flow_epic", "idats"), "/", blood.pd$Array_well))
blood$meta <- blood$meta %>% left_join(blood.pd %>% rename(Array_well = "sample_id"))
#saveRDS(blood, here("data", "blood_flow_epic", paste0("ewastools_read_idat_output_annotated_", Sys.Date(), ".rda")))
```

No Illumina controls failed
```{r, eval = F}
ctrls <- control_metrics(blood)
blood$meta$flag_ctrlfail <- ewastools::sample_failure(ctrls)
table(blood$meta$flag_ctrlfail)
```

Not sure why this was done; skipping in favor on same pipeline used for other samples
```{r, eval = F}
na.indices <- which(blood$meta$Sex %>% is.na())
blood <- blood %>% detectionP()
blood <- drop_samples(blood, j = na.indices)
male.indices <- which(blood$meta$Sex == "M")
female.indices <- which(blood$meta$Sex == "F")
eval_detP_cutoffs(blood.test, males = male.indices, female = female.indices)
```

# FlowSorted.Blood.EPIC QC w/ minfi

```{r}
rge <- read.metharray.exp(targets = blood.pd, extended = T)
```

```{r}
rge
```

```{r}
clin.covariates <- c("Sex", "CellType", "Sentrix_ID", "Sentrix_Position")
```

Create MethylSet
```{r}
rawMSet <- preprocessRaw(rge)
rawMSet

#saveRDS(rawMSet, file = here("data", "kc", paste0("methylSet_raw_", Sys.Date(), ".rda")))
```

```{r}
Meth <- getMeth(rawMSet)
Meth[1:5, 1:5]
```

```{r}
Unmeth <- getUnmeth(rawMSet) 
Unmeth[1:5, 1:5]
```

## Visualize Methylated and Unmethylated Intensities

Flag 1 low intensity sample
```{r}
pd <- pData(rge) %>% as.data.frame

# Overall intensity: M vs. U
pd$MQC <- log2(colMedians(Meth))
pd$UQC <- log2(colMedians(Unmeth))
```

Graphing function, taken from EGESS module 2
```{r}
intenseplot <- function(col = ""){
  ggplot(pd, aes_string(x = "UQC", y = "MQC", color = col)) + 
    geom_point() + 
    #coord_cartesian(xlim = c(11, 13), ylim = c (11, 13)) +
    labs(x = "Log2 Median Unmethylated Intensity", y = "Log2 Median Methylated intensity",
         color = sprintf("Sample %s", col)) + #sprintf() %s means 'insert a character string here' and col is the R object with the corresponding string
    theme_bw() +
    ggtitle(col)
}
```

```{r}
map(clin.covariates, intenseplot)
```

Any low intensity samples?
```{r}
# Drop (or if really small sample: watch out for): Samples with UQC<11 & MQC<11
# Note the cutoff value (here, 11) would depend on your data and array (EPIC/450k)
sum(pd$UQC < 11)
```

```{r}
sum(pd$MQC < 11)
```

```{r}
rm(Meth, Unmeth) # Clean up our coding environment
```

```{r}
pd <-
  pd %>%
  mutate(flag_lowintensity = ifelse(
    UQC < 11 & MQC < 11,
    TRUE,
    FALSE
  ))
```

```{r}
densityPlot(rawMSet, sampGroups = pd$CellType)
```

```{r}
densityBeanPlot(rawMSet, sampGroups = pd$CellType)
```

## Check sex
```{r}
grs <- rawMSet %>% mapToGenome() %>% ratioConvert()
```

```{r}
sex.info <- getSex(grs)

pd$predictedSex <- sex.info$predictedSex
pd$xMed <- sex.info$xMed
pd$yMed <- sex.info$yMed
```

Predicted sex matches up with annotated sex
```{r}
ggplot(data = pd, mapping = aes(x = xMed, y = yMed, color = predictedSex, fill = predictedSex)) + geom_point()
ggplot(data = pd, mapping = aes(x = xMed, y = yMed, color = Sex, fill = Sex)) + geom_point()
```

```{r}
rm(grs)
rm(rawMSet)
```


## Genotype

15 of 49 are outliers
```{r}
#snps <- getSnpInfo(grs)
snps.beta <- getSnpBeta(rge)
genotypes <- call_genotypes(snps.beta, learn = F)
pd$outlier <- snp_outliers(genotypes)

(pd$outlier < -4) %>% summary
```

The only matches that come up are replicates and mixes, as expected
```{r}
check_snp_agreement(genotypes, donor_ids = pd$Subject.ID, pd$Sample_Name)
```


## NBeads

```{r}
nbeads <- wateRmelon::beadcount(rge)

beadmean <- data.frame(bmean = rowMeans(nbeads, na.rm = T))

ggplot(beadmean, aes(x = bmean)) +
  geom_histogram(bins = 200, color = "black", fill = "grey") +
  coord_cartesian(xlim = c(0, 25)) +
  labs(x = "Probe mean beadcount") +
  theme_bw()
```

8 with < 5; 8 na
```{r}
summary(beadmean < 5)
```

```{r}
na.idx <- is.na(beadmean)
```

Start with 866091 sites
```{r}
ms.noob <- preprocessNoob(rge)
ms.noob
#saveRDS(ms.noob, here("data", "Campbell_Salas_Yuan_raw_ms_noob_2022-11-01.rda"))
```

The 8 sites with missing beadmean
```{r}
names(ms.noob[na.idx, ])
manifest %>% filter(probe_id %in% names(ms.noob[na.idx,]))
```

Drop the 16 probes with low bead count or missing
```{r}
keep.bead.idx <- (beadmean$bmean >= 5 & is.na(beadmean$bmean) == F)
ms.noob.nbead <- ms.noob[keep.bead.idx, ]
ms.noob.nbead
```

## detectionP
SNPs already get dropped by preprocessNoob it appears
```{r}
detectionP <- detectionP.minfi(rge)
detectionP <- detectionP[keep.bead.idx, ]
```

```{r}
detprobe <- data.frame(pmean = rowMeans(detectionP, na.rm = T)) 
probecut <- 0.01
```

```{r}
ggplot(detprobe, aes(x = pmean)) +
  geom_histogram(color = "black", fill = "grey", bins = 1000) +
  coord_cartesian(xlim = c(0, 0.08), ylim = c(0, 5e4)) +
  geom_vline(xintercept = probecut, color = "red", linetype = "dashed") +
  theme_bw() +
  labs(x = "Mean detectionP per probe", y = "Count") +
  ggtitle("All Samples")
```

 5594 probes have average detP > 0.01
```{r}
table(detprobe$pmean > 0.01)
```

Only 0.6% are over the threshold
```{r}
table(detprobe$pmean > 0.01) %>% '/'(nrow(detprobe)) %>% round(digits = 3)
```

Get the % detP failure rate (x/total number samples) for each probe
```{r}
prob.fail <- apply(X = detectionP, MARGIN = 1, FUN = function(x) (x > 0.01) %>% as.numeric() %>% mean)
```

4438 probes had greater than 5% of samples failure rate for detP
```{r}
probes.to.drop.detp <- prob.fail > 0.05
summary(probes.to.drop.detp)
```

Subset noob-normalized MethylSet to drop failed detP probes
```{r}
ms.noob.nbead.detP <- ms.noob.nbead[!probes.to.drop.detp,]
```

## Cross-reactive probes

Zhou et al., 2017 Alternative resource:"Comprehensive characterization, annotation and innovative use of Infinium DNA methylation BeadChip probes" {
Three groups of probes should generally be filtered out from Infinium microarray analyses: (i) probes with internal SNPs close to the 3΄ end of the probe (Group 1); (ii) probes with non-unique mapping to the bisulfite-converted genome (Group 2); and (iii) probes with off-target hybridization due to partial overlap with non-unique elements (Group 3). } http://zwdzwd.github.io/InfiniumAnnotation#current

Read in Pidsley et al., 2016 or Zhou, 2017 (will use) cross-reactive probes
```{r}
#cross.reactive <- read_csv(here("data" , "pidsley_cross-reactive_probes.csv"))
#MASK_general contains generally advised probes for dropping
cross.reactive <- read_tsv(here("data", "EPIC.hg19.manifest.tsv")) 
```

Get ProbeIDs to drop (~96k probes)
```{r}
cross.reactive.probes <- 
  cross.reactive %>%
  filter(MASK_general == T) %>%
  pull(probeID)
```

Drop cross-reactive probes from beta matrix
```{r}
ms.noob.nbead.detP.xreac <- ms.noob.nbead.detP[!rownames(ms.noob.nbead.detP) %in% cross.reactive.probes,]
```

97157 cross-reactive probes dropped
```{r}
#paste0(dim(ms.noob.nbead.detP)[[1]] - dim(ms.noob.nbead.detP.xreac)[[1]], " cross-reactive probes (Pidsley et al., 2016) dropped")
dim(ms.noob.nbead.detP.xreac)

paste0(dim(ms.noob.nbead.detP)[[1]] - dim(ms.noob.nbead.detP.xreac)[[1]], " cross-reactive probes (Zhou et al., 2017) dropped")
dim(ms.noob.nbead.detP.xreac)
```

## Drop sex-linked probes

Load ewastools manifest
```{r}
manifest = data.table::copy(ewastools:::manifest_epic)
```

Drop 16782 sex-linked probes
```{r}
chrXY.to.drop = manifest[chr%in%c("X","Y"),probe_id]
length(chrXY.to.drop)
ms.noob.nbead.detP.xreac.auto <- ms.noob.nbead.detP.xreac[!rownames(ms.noob.nbead.detP.xreac) %in% chrXY.to.drop,]

paste0(dim(ms.noob.nbead.detP.xreac)[[1]] - dim(ms.noob.nbead.detP.xreac.auto)[[1]], " sex-linked probes dropped")
dim(ms.noob.nbead.detP.xreac.auto)
```

## BMIQ

Convert to beta 
```{r}
beta <- getBeta(ms.noob.nbead.detP.xreac.auto)
```

BMIQ norm
```{r}
beta.bmiq <- champ.norm(beta, resultsDir = here("data", "analytic", "blood", "epic"), plotBMIQ = T, arraytype = "EPIC", cores = 8, method = "BMIQ")
rownames(beta.bmiq) <- rownames(beta)
colnames(beta.bmiq) <- colnames(beta)
```

```{r}
keep.ids <- blood.pd %>%
  filter(Subject.ID != "Th2535-2") %>%
  filter(CellType != "MIX") %>%
  pull(Array_well)
blood.pd.keep <- blood.pd %>%
  filter(Array_well %in% keep.ids)
beta.bmiq.keep <- beta.bmiq[,colnames(beta.bmiq) %in% keep.ids]
```

### Reformat to match other datasets

```{r}
blood.pd.keep.subset <-
  blood.pd.keep %>%
  mutate(Study = "Salas") %>%
  dplyr::select("Sample.ID" = Sample_Name,
                CellType,
                Study,
                Sex,
                "Case_ID" = Subject.ID,
                Sentrix_ID,
                Sentrix_Position
                ) %>%
  mutate_if(is.factor, as.character)
```


## Save result
747698 sites for analysis across 36 samples
```{r}
#saveRDS(beta.bmiq.keep, file = here("data", "analytic", "blood", paste0("qc_beta_bmiq_blood_2023-11-16.rda")))
#saveRDS(blood.pd.keep.subset, file = here("data", "analytic", "blood", paste0("qc_pd_blood_2023-11-16.rda")))
```
