---
title: "Merge raw and QC placental cell type"
author: "Kyle Abraham Campbell"
date: '2022-09-22'
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# CRAN packages
library(BiocManager)
library(corrplot)
library(data.table)
library(FactoMineR)
library(factoextra)
library(GGally)
library(GEOquery)
library(ggrepel)
library(here)
library(readxl)
library(svd)
library(tidymodels)
library(tidyverse)
library(viridis)

# BioConductor packages
library(ChAMP)
library(ExperimentHub)
library(FlowSorted.CordBlood.450k)
library(illuminaio)
library(IlluminaHumanMethylationEPICmanifest)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(minfi)
#library(wateRmelon)


# Github packages
#devtools::install_github("hhhh5/ewastools@master")
library(ewastools)
#devtools::install_github('wvictor14/planet')
library(planet)
#remotes::install_github("immunomethylomics/FlowSorted.Blood.EPIC")
library(FlowSorted.Blood.EPIC)
#remotes::install_github("immunomethylomics/FlowSorted.BloodExtended.EPIC")
library(FlowSorted.BloodExtended.EPIC)
```

```{r}
# Read the RGSet
cord <-
    libraryDataGet('FlowSorted.CordBloodCombined.450k')
cord.pd <- pData(cord) %>% as.data.frame

# Get deGoede indices and subset accordingly
study.idx <- cord.pd$Study %in% c("deGoede")
cord.pd <- cord.pd[study.idx,]
cord <- cord[, study.idx]

# Reformat cord.pd
# Reminder on rename syntax: new_name = old_name
cord.pd.x <- cord.pd %>%
  dplyr::rename(
    Sample.ID = "Sample_Name",
    Case_ID = "Subject.ID",
    Sentrix_ID = "Slide",
    Sentrix_Position = "Array"
  ) %>%
  mutate(Sample.ID = ifelse(
    Study == "deGoede",
    paste0(Case_ID, "_", CellType_original),
    Sample.ID
  ))

# Reassign pd on RGset
pData(cord) <- as(cord.pd.x, "DataFrame")

#cord.nrbc.only <- cord[,cord$CellType == "nRBC"]
#cord.nrbc.only.noob <- preprocessNoob(cord.nrbc.only)
# Save a raw RGSet w/ just nRBC samples
#saveRDS(cord.nrbc.only, here("data", "cord", "raw_rgset_cord_nrbc_only.rda"))
```

```{r}
rge <- cord
pd <- cord.pd.x
#rm(cord)
#rm(cord.pd)
#rm(cord.nrbc.only)
```

```{r}
# Note, not actually RGSetExtended, just keeping name the same for ease of porting code
#rge <- readRDS(here("data", "cord", "cord_nrbc_formatted_rgset_10-17-2022.rda"))
#pd <- pData(rge) %>% as.data.frame()
```

```{r}
manifest = data.table::copy(ewastools:::manifest_450K)
```

```{r}
clin.covariates <- c("Sex", "CellType", "Case_ID", "Study")
```

Create MethylSet
```{r}
rawMSet <- preprocessRaw(rge)
rawMSet
```

```{r}
Meth <- getMeth(rawMSet)
Meth[1:5, 1:5]
```

```{r}
Unmeth <- getUnmeth(rawMSet) 
Unmeth[1:5, 1:5]
```

## Visualize Methylated and Unmethylated Intensities

Flag 0 low intensity sample
```{r}
pd <- pData(rge) %>% as.data.frame()

# Overall intensity: M vs. U
pd$MQC <- log2(colMedians(Meth))
pd$UQC <- log2(colMedians(Unmeth))
```

Graphing function, taken from EGESS module 2
```{r}
intenseplot <- function(col = ""){
  ggplot(pd, aes_string(x = "UQC", y = "MQC", color = col)) + 
    geom_point() + 
    #coord_cartesian(xlim = c(11, 13), ylim = c (11, 13)) +
    labs(x = "Log2 Median Unmethylated Intensity", y = "Log2 Median Methylated intensity",
         color = sprintf("Sample %s", col)) + #sprintf() %s means 'insert a character string here' and col is the R object with the corresponding string
    theme_bw() +
    ggtitle(col)
}
```

```{r}
map(clin.covariates, intenseplot)
```

Any low intensity samples?
```{r}
# Drop (or if really small sample: watch out for): Samples with UQC<11 & MQC<11
# Note the cutoff value (here, 11) would depend on your data and array (EPIC/450k)
sum(pd$UQC < 11)
```

```{r}
sum(pd$MQC < 11)
```

```{r}
rm(Meth, Unmeth) # Clean up our coding environment
```

```{r}
pd <-
  pd %>%
  mutate(flag_lowintensity = ifelse(
    UQC < 11 & MQC < 11,
    TRUE,
    FALSE
  )) %>%
  mutate(CellType = factor(CellType))
```

```{r}
densityPlot(rawMSet, sampGroups = pd$CellType, legend = F)
```

```{r}
densityBeanPlot(rawMSet, sampGroups = pd$CellType)
```

## Check sex
```{r}
grs <- rawMSet %>% mapToGenome() %>% ratioConvert()
```

```{r}
sex.info <- getSex(grs)

pd$predictedSex <- sex.info$predictedSex
pd$xMed <- sex.info$xMed
pd$yMed <- sex.info$yMed
```

Predicted sex matches up with annotated sex
```{r}
ggplot(data = pd, mapping = aes(x = xMed, y = yMed, color = predictedSex, fill = predictedSex)) + geom_point()
ggplot(data = pd, mapping = aes(x = xMed, y = yMed, color = Sex, fill = Sex)) + geom_point()
```


```{r}
pd %>%
  filter(predictedSex != Sex)
```

## Genotype

0 of 52 are outliers but no disagreements
```{r}
#snps <- getSnpInfo(grs)
snps.beta <- getSnpBeta(rge)
genotypes <- call_genotypes(snps.beta, learn = F)
pd$outlier <- snp_outliers(genotypes)

(pd$outlier < -4) %>% summary
```

No concerning conflicts found; genotype agreement was a little low ~.86 for two VY samples; the replicate sample from salas came up as agreement, as expected
```{r}
check_snp_agreement(genotypes, donor_ids = pd$Case_ID, sample_ids = pd$Sample.ID)
```

## NBeads
We don't have information on beads for the DeGeode samples, so we won't filter on those

Start with 453,093 sites
```{r}
ms.noob <- preprocessNoob(rge)
ms.noob
```

## detectionP
SNPs already get dropped by preprocessNoob it appears
```{r}
detectionP <- detectionP.minfi(rge)
```

```{r}
detprobe <- data.frame(pmean = rowMeans(detectionP, na.rm = T)) 
probecut <- 0.01
```

```{r}
ggplot(detprobe, aes(x = pmean)) +
  geom_histogram(color = "black", fill = "grey", bins = 1000) +
  coord_cartesian(xlim = c(0, 0.08), ylim = c(0, 5e4)) +
  geom_vline(xintercept = probecut, color = "red", linetype = "dashed") +
  theme_bw() +
  labs(x = "Mean detectionP per probe", y = "Count") +
  ggtitle("All Samples")
```

10955 probes have average detP > 0.01 w/ only nRBC samples
```{r}
table(detprobe$pmean > 0.01)
```

2.4% are over the threshold w/ only nRBC samples
```{r}
table(detprobe$pmean > 0.01) %>% '/'(nrow(detprobe)) %>% round(digits = 3)
```

Get the % detP failure rate (x/total number samples) for each probe
```{r}
prob.fail <- apply(X = detectionP, MARGIN = 1, FUN = function(x) (x > 0.01) %>% as.numeric() %>% mean)
```

21219 probes had greater than 5% of samples failure rate for detP
```{r}
probes.to.drop.detp <- prob.fail > 0.05
summary(probes.to.drop.detp)
```

Subset noob-normalized MethylSet to drop failed detP probes
```{r}
ms.noob.detP <- ms.noob[!probes.to.drop.detp,]
```

## Cross-reactive probes

Zhou et al., 2017 Alternative resource:"Comprehensive characterization, annotation and innovative use of Infinium DNA methylation BeadChip probes" {
Three groups of probes should generally be filtered out from Infinium microarray analyses: (i) probes with internal SNPs close to the 3΄ end of the probe (Group 1); (ii) probes with non-unique mapping to the bisulfite-converted genome (Group 2); and (iii) probes with off-target hybridization due to partial overlap with non-unique elements (Group 3). } http://zwdzwd.github.io/InfiniumAnnotation#current

Read in Pidsley et al., 2016 or Zhou, 2017 (will use) cross-reactive probes
```{r}
#cross.reactive <- read_csv(here("data" , "pidsley_cross-reactive_probes.csv"))
#MASK_general contains generally advised probes for dropping
cross.reactive <- read_tsv(here("data", "HM450.hg19.manifest.tsv")) 
```

Get ProbeIDs to drop (~60k probes)
```{r}
cross.reactive.probes <- 
  cross.reactive %>%
  filter(MASK_general == T) %>%
  pull(probeID)
length(cross.reactive.probes)
```

Drop cross-reactive probes from beta matrix
```{r}
ms.noob.detP.xreac <- ms.noob.detP[!rownames(ms.noob.detP) %in% cross.reactive.probes,]
```

51884 cross-reactive probes dropped
```{r}
#paste0(dim(ms.noob.detP)[[1]] - dim(ms.noob.detP.xreac)[[1]], " cross-reactive probes (Pidsley et al., 2016) dropped")
dim(ms.noob.detP.xreac)

paste0(dim(ms.noob.detP)[[1]] - dim(ms.noob.detP.xreac)[[1]], " cross-reactive probes (Zhou et al., 2017) dropped")
dim(ms.noob.detP.xreac)
```

## Drop sex-linked probes

Load ewastools manifest
```{r}
manifest = data.table::copy(ewastools:::manifest_450K)
```

Drop 8705 sex-linked probes
```{r}
chrXY.to.drop = manifest[chr%in%c("X","Y"),probe_id]
length(chrXY.to.drop)
ms.noob.detP.xreac.auto <- ms.noob.detP.xreac[!rownames(ms.noob.detP.xreac) %in% chrXY.to.drop,]

paste0(dim(ms.noob.detP.xreac)[[1]] - dim(ms.noob.detP.xreac.auto)[[1]], " sex-linked probes dropped")
dim(ms.noob.detP.xreac.auto)
```

## BMIQ normalization w/ CHAMP

```{r}
beta = ms.noob.detP.xreac.auto %>% getBeta()
```

Interactive QC plots from ChAMP
```{r, eval = F}
QC.GUI(beta=beta, arraytype="450K", pheno=pd$Study)
```

Write QC plots to directory with ChAMP
```{r, eval = F}
champ.QC(beta=beta, pheno=pd$Study, resultsDir = here("results", "nrbc_bmiq_deGoede"), dendrogram = T, mdsPlot = T)
```

```{r}
beta.bmiq <- champ.norm(beta, resultsDir = here("results", "nrbc_bmiq_deGoede"), plotBMIQ = T, arraytype = "450K", cores = 8)
rownames(beta.bmiq) <- rownames(beta)
colnames(beta.bmiq) <- colnames(beta)
```

## Pull only the nRBC profiles
```{r}
nrbc.idx <- which(pd$CellType == "nRBC")
nrbc.idx

## Drop non-nRBC profiles from RGSet and pd 
nrbc.pd <- pd[nrbc.idx, ]
nrbc <- ms.noob.detP.xreac.auto[, nrbc.idx]
pData(nrbc) <- as(nrbc.pd, "DataFrame")

#saveRDS(nrbc.pd, file = here("data", "analytic", "nrbc", "deGoede_nrbc_pd_2023-11-16.rda"))
```

```{r}
nrbc
#saveRDS(nrbc %>% getBeta(), file = here("data", "cord", "beta_noob_nrbc_qc_w_deGoede_deGoede_2022-10-20.rda"))
```

## Save result
371285 sites for analysis across 7 samples
```{r}
beta.bmiq.nrbc <- beta.bmiq[, nrbc.idx]
dim(beta.bmiq.nrbc)
#saveRDS(beta.bmiq.nrbc, file = here("data", "analytic", "nrbc", "beta_noob_bmiq_nrbc_qc_deGoede_2023-11-15.rda"))
```


```{r}
#saveRDS(ms.noob.detP.xreac.auto, file = here("data", "cord", paste0("qc_methylset_noob_cord_", Sys.Date(), ".rda")))
#saveRDS(ms.noob.detP.xreac.auto %>% getBeta(), file = here("data", "cord", paste0("qc_methylset_noob_beta_cord_", Sys.Date(), ".rda")))
#saveRDS(beta.bmiq, file = here("data", "cord", paste0("qc_methylset_noob_bmiq_cord_", Sys.Date(), ".rda")))
```