---
title: "GSE232778_compare_decon_methods"
author: "Kyle Abraham Campbell"
date: "2023-11-01"
output: html_document
---

Based on methods presented in:
Reliability of a novel approach for reference-based cell type estimation in human placental DNA methylation studies

```{r setup, include=FALSE}
#library(devtools)
library(corrplot)
library(devtools)
library(EpiDISH)
library(genefilter)
library(ggcorrplot)
library(here)
library(factoextra)
library(minfi)
library(openxlsx)
library(pheatmap)
library(planet)
library(skimr)
library(tidyverse)
# For pickCompProbes local implementation
source(here("scripts", "analytic", "00_custom_helper_scripts", "10_custom_functions.R"))

#install_github("Github-MS/xvalglms")
library(xvalglms)

knitr::opts_chunk$set(echo = TRUE)
```

## Load data
204 whole tissue placental samples
```{r}
gse <- read_csv(here("data", "gse232778", "GSE232778_matrixtable_processed_betas.csv")) %>% column_to_rownames(var = "ID_REF")
gse.pd <- read_csv(here("data", "gse232778", "GSE232778_pdata.csv"))

keep <- gse.pd %>% filter(`characteristics:Replicate_Removed` == F) %>% pull(`characteristics:Sentrix`)

gse.pd <- gse.pd %>% filter(`characteristics:Replicate_Removed` == F)
gse <- gse[,colnames(gse) %in% keep]

data("plCellCpGsThird")
```


## Examing Yuan cell type reference probes

All Yuan probes are on 450k
```{r, eval = F}
cpgs.450k <- IlluminaHumanMethylation450kanno.ilmn12.hg19::Manifest %>% rownames
# Load Planet CpGs
data("plCellCpGsThird")
rownames(plCellCpGsThird) %in% cpgs.450k %>% summary
# Most of Yuan probes are NOT in written-up Campbell reference, even though there are twice as many
rownames(plCellCpGsThird) %in% rownames(celltypes.ref) %>% summary
```


## Deconvolution estimates from Campbell reference
### Load
```{r}
#pd <- readRDS(here("data", "analytic", "concat", "pd_2023-11-16.rda")) %>% mutate(CellType = factor(CellType))
pd <- readRDS(here("data", "analytic", "concat", "pd_2023-11-30.rda")) %>% mutate(CellType = factor(CellType))
beta <- readRDS(here("data", "analytic", "concat", "beta_2023-11-16.rda"))
#beta <- beta[,colnames(beta) %in% pd$Sample.ID]
```

### pickCompProbes
```{r}
p <- beta[,colnames(beta) %in% (pd %>% filter(CellType != "Villi") %>% pull(Sample.ID))]

p.villi <- beta[,colnames(beta) %in% (pd %>% filter(CellType == "Villi") %>% pull(Sample.ID))]

pd.fn <- pd %>% filter(CellType != "Villi")
p.pd <- pd %>% filter(CellType != "Villi")
pd.villi <- pd %>% filter(CellType == "Villi")

res <- pickCompProbes_kc(
  p = beta[,colnames(beta) %in% (pd %>% filter(CellType != "Villi") %>% pull(Sample.ID))],
  pd.fn = pd %>% filter(CellType != "Villi"),
  numProbes = 30
  )
```

Export reference package datasets
```{r, eval = F}
save(p.villi, file = here("data", "analytic", "reference_package", "p_villi.rda"))
save(pd.villi, file = here("data", "analytic", "reference_package", "pd_villi.rda"))
save(p, file = here("data", "analytic", "reference_package", "p.rda"))
save(p.pd, file = here("data", "analytic", "reference_package", "pd.rda"))

ref <- res$coefEsts[, colnames(res$coefEsts) != "Villi"]
ref.df <- ref %>% as.data.frame()

save(ref, file = here("data", "analytic", "reference_package", "ref.rda"))
#save(ref, file = here("ref.rda"))
#load(here("data", "analytic", "reference_package", "ref.rda"))
write_csv(x = ref.df, here("ref.csv"))
```

### Deconvolution
```{r}
epidish.res <- epidish(beta.m = gse, ref.m = res$coefEsts[, colnames(res$coefEsts) != "Villi"], method = "RPC")
#head(epidish.res$estF)
cor(epidish.res$estF[,12], gse.pd$`characteristics:Syncytiotrophoblast`)

x <- epidish.res$estF %>% as.data.frame() %>% rownames_to_column(var = "characteristics:Sentrix")

gse.pd.x <- left_join(gse.pd, x)
```

## Create PCA of gse matrix
PC1: 15.4%
PC2: 5.9%
PC3: 3.5%
PC4: 3.2%
PC5: 2.0%
PC6: 1.9%
PC7: 1.6%
PC8: 1.4%
PC9: 1.1%
PC10: 1.1%
```{r}
# Perform PCA w/o scaling because beta values are all on same scale
res.pca <- prcomp(t(gse), scale = FALSE)
summary(res.pca)
fviz_eig(res.pca)
res.pca$x
res.pca.x <- res.pca$x %>% as.data.frame() %>% dplyr::select(PC1:PC10) %>% rownames_to_column(var = "characteristics:Sentrix")

pd.x.x <- gse.pd.x %>%
  left_join(res.pca.x) %>%
  dplyr::select(!c("characteristics:Sentrix_ID", "characteristics:AIMs_Coordinate_1", "characteristics:AIMs_Coordinate_2", "characteristics:AIMs_Coordinate_3", "characteristics:PlaNET_GA_RRPC", "characteristics:PlaNET_Prob_African", "characteristics:PlaNET_Prob_Asian", "characteristics:PlaNET_Prob_European")) 
pd.x.x$PC1 %>% summary()
```

### Reformat phenotype matrix

```{r}
colnames(pd.x.x) <- gsub(pattern = "characteristics:", replacement = "planet_", x = colnames(pd.x.x))
colnames(pd.x.x)

pd.x.x.num <- pd.x.x %>%
  dplyr::select(where(is.numeric)) %>%
  drop_na() %>% 
  dplyr::rename(
    Birthweight_g = planet_Birthweight_g,
    Gestational_age = planet_GA_Reported,
  ) %>%
  #dplyr::select(!c(PC7, PC8, PC9, PC10)) %>%
  dplyr::select(!c(Birthweight_g, Gestational_age, planet_Birthweight_SD)) %>%
  dplyr::rename(
    `PC1 (15.4%)` = PC1,
    `PC2 (5.9%)` = PC2,
    `PC3 (3.5%)` = PC3,
    `PC4 (3.2%)` = PC4,
    `PC5 (2.0%)` = PC5,
    `PC6 (1.9%)` = PC6,
    `PC7 (1.6%)` = PC7,
    `PC8 (1.4%)` = PC8,
    `PC9 (1.1%)` = PC9,
    `PC10 (1.1%)` = PC10
  )
colnames(pd.x.x.num)
```

## Check cor matrix
```{r}
# Example of what's used below
#pd.x.x %>% dplyr::select(where(is.numeric)) %>% dplyr::select(!c(nRBC, "characteristics:Sentrix_ID", "characteristics:AIMs_Coordinate_1", "characteristics:AIMs_Coordinate_2", "characteristics:AIMs_Coordinate_3", "characteristics:PlaNET_GA_RRPC", "characteristics:PlaNET_Prob_African", "characteristics:PlaNET_Prob_Asian", "characteristics:PlaNET_Prob_European"))

# Dropped 2 missing values for reported GA
M = cor(pd.x.x.num)
#ggcorrplot(M)
pmat <- cor_pmat(M)
ggcorrplot <- ggcorrplot(M, type = "lower", tl.cex = 7, p.mat = pmat)
ggcorrplot
#ggsave(plot = ggcorrplot, here("results", "analytic", "xval_glm", "corrplot.tiff"), dpi = 300)
```


## Setup models
Drop Syncytiotrophoblast from models to avoid collinearity issues
```{r}
#Vectorized

list.models =
  list(
    "PC1" = list(PC1 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC1 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    
    "PC2" = list(PC2 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC2 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC3" = list(PC3 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC3 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC4" = list(PC4 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC4 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC5" = list(PC5 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC5 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC6" = list(PC6 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC6 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC7" = list(PC7 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC7 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC8" = list(PC8 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC8 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC9" = list(PC9 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono  + nRBC,
    PC9 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`),
    "PC10" = list(PC10 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC10 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial` + `planet_nRBC`)
  )

#models =
#  list(
#    "Combined reference" = PC1 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + #CD4T + CD8T + Mono,
#    "planet alone" = PC1 ~ `characteristics:Cytotrophoblast` + `characteristics:Stromal` + #`characteristics:Hofbauer` + `characteristics:Endothelial`
#  )
```

Original
```{r, eval = F}
models =
  list(
    PC1 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC1 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC2 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC2 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC3 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC3 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC4 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC4 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC5 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC5 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC6 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC6 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC7 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC7 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC8 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC8 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC9 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC9 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`,
    PC10 ~ Cytotrophoblast + Hofbauer + Stromal + Endothelial + Neu + NK + Bcell + CD4T + CD8T + Mono,
    PC10 ~ `planet_Cytotrophoblast` + `planet_Stromal` + `planet_Hofbauer` + `planet_Endothelial`
  )
```

## Execute models
vectorized
```{r}
res.xval <-
  map(list.models, ~ xval.glm(
    data = pd.x.x,
    .x,
    glm.family = gaussian,
    folds = 10,
    repeats = 1000, #normally 200, we used 1000
    loss = NULL,
    numCore = NULL,
    plots = T,
    gray = F,
    seed = 5,
    showConsoleOutput = T
  ))
#saveRDS(res.xval, here("results", "analytic", "xval_glm", "xval_glm_vectorized_proposed_analytic.rds"))
#ggsave(here("results", "analytic", "figures", "figure_x_xval.glm.jpg"))
#ggsave(here("results", "analytic", "figures", "figure_x_xval.glm.jpeg"), dpi=300)

#res.xval$data
res.xval$PC1$box.plot
res.xval$PC2$box.plot
res.xval$PC3$box.plot
res.xval$PC4$box.plot
res.xval$PC5$box.plot
res.xval$PC6$box.plot
res.xval$PC7$box.plot
res.xval$PC8$box.plot
res.xval$PC9$box.plot
res.xval$PC10$box.plot

res.xval$PC1$RMSEP$PC <- "PC1 (15.4%)"
res.xval$PC2$RMSEP$PC <- "PC2 (5.9%)"
res.xval$PC3$RMSEP$PC <- "PC3 (3.5%)"
res.xval$PC4$RMSEP$PC <- "PC4 (3.2%)"
res.xval$PC5$RMSEP$PC <- "PC5 (2.0%)"
res.xval$PC6$RMSEP$PC <- "PC6 (1.9%)"
res.xval$PC7$RMSEP$PC <- "PC7 (1.6%)"
res.xval$PC8$RMSEP$PC <- "PC8 (1.4%)"
res.xval$PC9$RMSEP$PC <- "PC9 (1.1%)"


PC1 <- res.xval$PC1$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC1$wins[1], res.xval$PC1$wins[2]))
PC2 <- res.xval$PC2$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC2$wins[1], res.xval$PC2$wins[2]))
PC3 <- res.xval$PC3$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC3$wins[1], res.xval$PC3$wins[2]))
PC4 <- res.xval$PC4$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC4$wins[1], res.xval$PC4$wins[2]))
PC5 <- res.xval$PC5$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC5$wins[1], res.xval$PC5$wins[2]))
PC6 <- res.xval$PC6$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC6$wins[1], res.xval$PC6$wins[2]))
PC7 <- res.xval$PC7$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC7$wins[1], res.xval$PC7$wins[2]))
PC8 <- res.xval$PC8$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC8$wins[1], res.xval$PC8$wins[2]))
PC9 <- res.xval$PC9$RMSEP %>% mutate(Wins = ifelse(Model == 1, res.xval$PC9$wins[1], res.xval$PC9$wins[2]))

PC <- bind_rows(PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9) %>%
  mutate(Model = ifelse(Model == 1, "Expanded reference", "planet reference"))
PC$Model <- factor(PC$Model)


rmse_plot <- ggplot(data = PC, mapping = aes(x = PC, y = RMSEP, fill = Model)) +
  # Boxplot for RMSE values grouped by PC and Model
  geom_boxplot(position = position_dodge(width = 0.75), outlier.shape = NA, width = 0.7) +
  # Add jittered points for individual RMSE values
  geom_jitter(aes(color = Model), position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75), size = 2, alpha = 0.5) +
  theme_bw() +
  # Annotate the number of wins at the top of each boxplot
  stat_summary(aes(label = Wins), fun = max, geom = "text", 
               position = position_dodge(width = 0.75), vjust = -1, size = 3.5, color = "black") +
  # Customize the theme for better visualization
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "RMSEP by Model and Principal Component", 
       subtitle = "Wins of 1000 cross-validation models for each PC",
       x = "Principal Component (PC)", 
       y = "RMSEP",
       fill = "Model", 
       color = "Model") +
  scale_fill_manual(values = c("lightblue", "salmon")) +  # Custom colors
  scale_color_manual(values = c("lightblue", "salmon")) +
  ylim(0,12)

#ggsave(filename = here("results", "analytic", "xval_glm", "rmse_plot.tiff"), plot = rmse_plot)
```

Non-vectorized
```{r, eval = F}
res.xval.single <-
  xval.glm(
    data = pd.x.x,
    models = models,
    glm.family = gaussian,
    folds = 10,
    repeats = 100, #normally 200, we used 1000
    loss = NULL,
    numCore = NULL,
    plots = T,
    gray = F,
    seed = 5,
    showConsoleOutput = T
  )
#res.xval$data
```

Summary statistics
```{r}
#PC %>% group_by(PC) %>% group_by(Model) %>% tbl_summary()

rmse <- res.xval$RMSEP
write.xlsx(rmse, here("results", "analytic", "figures", "figure_x_xval.glm.xlsx"))
rmse %>% group_by(Model) %>% summarise(Mean = mean(RMSEP), SD = sd(RMSEP))
```

Test for differences
```{r}
x1 <- rmse %>% filter(Model == 1) %>% pull(RMSEP)
x2 <- rmse %>% filter(Model == 2) %>% pull(RMSEP)
wilcox.test(x1, x2)
t.test(x1, x2)
```

```{r}
hist(x1)
hist(x2)
```

Test for differences across models
```{r}
# If you want to use a non-parametric test (e.g., Wilcoxon rank-sum test) instead of t-test:
test_results <- list()

for (pc in unique(PC$PC)) {
  data_subset <- subset(PC, PC == pc)
  wilcox_test_result <- wilcox.test(RMSEP ~ Model, data = data_subset)
  test_results[[pc]] <- wilcox_test_result
}

# View the results
test_results
```


```{r}
res.xval$box.plot
res.xval$box.plot + scale_x_discrete(labels = c("Combined reference", "planet alone"))
```

```{r}
rmse %>%
  mutate(Model = ifelse(
    Model == 1,
    "Combined reference",
    "planet alone"
  ))


p <- ggplot_build(res.xval$box.plot)
p$data[[1]]$group[1] = "Combined reference"
p$data[[1]]$group[2] = "planet alone"

p$data[[2]] <- p$data[[2]] %>% mutate(group = ifelse(
    group == 1,
    "Combined reference",
    "planet alone"
  ))

q <- ggplot_gtable(p)
plot(q)
```

Some plotting of relative estimates 

```{r}
ggplot(dat = pd.x.x, mapping = aes(x = Cytotrophoblast, y = `characteristics:Cytotrophoblast`)) + geom_point()
skim(pd.x.x$Cytotrophoblast)
skim(pd.x.x$`characteristics:Cytotrophoblast`)

z <- pd.x.x %>% dplyr::select(`characteristics:Syncytiotrophoblast`:nRBC) %>% skim
z.x <- z %>% dplyr::select(!c(skim_type, n_missing, complete_rate, character.min, character.max, character.empty, character.n_unique, character.whitespace)) %>% filter(!skim_variable %in% c("characteristics:Basename", "characteristics:molecule")) %>% mutate(across(where(is.numeric), round, digits = 2))
```
